---
title: "UseR group data visualization contest"
output: html_notebook
---



```{r}
library(tidyverse)
library(readxl)
library(gghighlight)
```


# LOAD DATASET
```{r}
mortality_tbl <- read_excel("global_mortality.xlsx")
```


```{r}
mortality_tbl
```

# DATA WRANGLING
# pivot_longer
```{r}
mortality_tbl_long <- mortality_tbl %>% 
  
  # pivot tibble to long format
  pivot_longer(ends_with("(%)"), names_to = "cause", values_to = "proportion")

mortality_tbl_long

```


# remove "(%)" from the end of the strings in the cause columns and trim strings to remove space at the end
```{r}
mortality_tbl_long <- mortality_tbl_long %>% 
  mutate(cause = str_extract(mortality_tbl_long$cause, "[^(%)]+")) %>% 
  mutate(cause = cause %>% str_trim())
```





# list causes
```{r}
mortality_tbl_long %>% 
  
  distinct(cause)
```

# list countries/regions
```{r}
mortality_tbl_long %>% 
  
  distinct(country)
```






# SELECTED CAUSES OF DEATH VS. THE REST OF DISEASES BY COUNTRY


```{r}
selection_cause <- c("Cancers")

mortality_tbl_long %>% 
  mutate(plot_color = case_when(cause %in% selection_cause ~ "green",
                                TRUE ~ "lightgrey")
         )
  
```


```{r}
mortality_tbl_long %>% 
  
  filter(country == "United States") %>% 
  
  ggplot(aes(x = year, y = proportion)) +
  
  geom_line(aes(group = cause, color = cause)) +
  gghighlight(cause %in% selection_cause,
              unhighlighted_params = list(size = 1, colour = alpha("grey", 0.2))) +
  
  theme_light() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  
  scale_color_viridis_d()
```


# function for between causes comparison by country
```{r}
create_btw_cause_by_country <- function(cause_of_death, ctr) {
  
  mortality_tbl_long %>% 
  
  filter(country == ctr) %>% 
  
  ggplot(aes(x = year, y = proportion)) +
  
  geom_line(aes(group = cause, color = cause)) +
  gghighlight(cause %in% cause_of_death,
              unhighlighted_params = list(size = 1, colour = alpha("grey", 0.2))) +
  
  theme_light() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  
  scale_color_viridis_d()
  
}
```

```{r}
create_btw_cause_by_country(c("Cancers", "Road accidents", "Homicide", "Terrorism"), c("Hungary"))
```







# SELECTED COUNTRIES VS. REST OF THE WORLD BY CAUSE

```{r}
selection <- c("Hungary", "United States", "Peru", "France", "Slovakia", "Poland", "Bhutan")

mortality_tbl_long_by_country <- mortality_tbl_long %>% 
  mutate(plot_color = case_when(country %in% selection ~ "green",
                                TRUE ~ "lightgrey")
         )

mortality_tbl_long_by_country

```


```{r}
mortality_tbl_long %>% 
  
  filter(cause == "Diabetes") %>% 
  
  
  ggplot(aes(x = year, y = proportion, color = country), size = 2) +
  
  geom_line() +
  gghighlight(country %in% selection,
              unhighlighted_params = list(size = 1, colour = alpha("lightgrey", 0.12))) +
  
  theme_light() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  
  scale_color_viridis_d()
```

# function for between country comparison by cause
```{r}
create_btw_country_by_cause <- function(countries, cause_of_death) {
  
mortality_tbl_long %>% 
  
  filter(cause == cause_of_death) %>% 
  
  
  ggplot(aes(x = year, y = proportion, color = country), size = 2) +
  
  geom_line() +
  gghighlight(country %in% countries,
              unhighlighted_params = list(size = 1, colour = alpha("lightgrey", 0.12))) +
  
  theme_light() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  
  scale_color_viridis_d()  
  
  
}


```


```{r}
create_btw_country_by_cause(c("Hungary", "United States", "Canada", "Peru"), c("HIV/AIDS"))
```





# COMPARISON BETWEEN SELECTED COUNTRIES BY CAUSE - SMALL MULTIPLES



# small multiples plot
```{r}
mortality_tbl_long %>% 
  
  filter(country %in% c("United States", "Hungary", "Germany", "Canada", "United Kingdom", "Croatia")) %>%
  filter(cause == "Neonatal deaths") %>% 
  
  ggplot(aes(x = year, y = proportion)) +
  
  geom_line(aes(group = cause), color ="lightgrey", size = 0.8) +
  
  facet_wrap(~ country, nrow = 2) +
  
  theme_tq() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  coord_cartesian(ylim = c(0, NA))
  
```
# function for small multiples plot
```{r}
create_small_multiples <- function(countries, cause_of_death) {
  
  mortality_tbl_long %>% 
  
  filter(country %in% countries) %>%
  filter(cause == cause_of_death) %>% 
  
  ggplot(aes(x = year, y = proportion)) +
  
  geom_line(aes(group = cause), color ="lightgrey", size = 0.8) +
  
  facet_wrap(~ country, nrow = 2) +
  
  theme_tq() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  coord_cartesian(ylim = c(0, NA))
  
}
```

```{r}
create_small_multiples(c("Hungary", "Spain", "Czech Republic", "United States", "Mexico", "Philippines", "Slovakia", "Poland"), "Neonatal deaths")
```

